---
output: github_document
---


<!-- README.md is generated from README.Rmd. Please edit that file -->


```{r setup, echo = FALSE, message = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-"
)
cvr <- covr::percent_coverage(covr::package_coverage())
coverage <- sprintf(
  "https://img.shields.io/badge/coverage-%0.1f%%25-%s.svg?style=flat&logo=codecov",
  cvr, ifelse(cvr > 75, "success", ifelse(cvr > 50, "yellow", "critical"))
)
#lints <- length(lintr::lint_dir("R"))
lints <- 0
lintage <- sprintf(
  "https://img.shields.io/badge/lints-%s-%s.svg?style=flat&logo=gitlab",
  lints, ifelse(lints < 25, "success", ifelse(lints < 50, "yellow", "critical"))
)
```


# The `libml` package

<!-- badges: start -->
[![License: GPL-3](https://img.shields.io/badge/License-GPL3-blue.svg)](https://www.gnu.org/licenses/gpl-3.0)
<!-- badges: end -->




## Overview

The `libml` package contains general functions 
necessary for the analysis of biomarker data via statistical and 
machine learning techniques, calculating KS distances, random forest,
k-fold cross-validation, classification performance, and ROC plotting.


----------------


## Custom Installation

```{r install-git, eval = FALSE}
remotes::install_github("libml")
```


----------------


## Usage

To load `libml` simply make a call to `library()` as usual:

```{r load, eval = TRUE}
library(libml)
```


## Help summary of the package

```{r help, eval = FALSE}
library(help = libml)
```


## Notes about the package

* The `libml` package is easy to use.
* There is some cool stuff we can do.



-------------

#### LICENSE


