---
title: "Introduction to libml"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to libml}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
library(libml)
library(randomForest)
library(ggplot2)
options(width = 100)
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>"
)
set.seed(101)
```



The `libml` package contains general functions 
necessary for the analysis of biomarker data via statistical and 
machine learning techniques, calculating KS distances, 
k-fold cross-validation, classification performance metrics, and ROC plotting.


---------------------


## Introduction

The `libml` package provides functions to create training datasets, 
perform univariate tests (via the `calc.*` function family,
train classifiers, and plot ROC curves. 
The models types supported in this package are primarily geared toward:

* Naive Bayes
* Random forest
* Logistic regression

There is also limited support for `k`-nearest neighbors models.
With the exception of "robust" Naive Bayes, these models are created using 
the associated R packages directly. The `libml` package provides 
wrappers for performing cross validation, making predictions, and creating ROC curves.

This package is biased toward *binary* classification and as a result many 
of the functions assume only two classes exist. To keep track of which 
class is the positive class, a `tibble` called a `tr_data` object is defined
for use in this package. A `tr_data` object is simply a `soma.adat` using
loaded using `SomaReadr::read.adat()`, with a "Response" column as the final
column. The "Response" column is a `factor` class, ordered so that the 
*positive* class is the **second** label. These `tr_data` data frames can be
created using the `create_train()` function.

ROC curves and other related functions all on a standard input consisting of:

* **truth**: a vector of actual class labels
* **predicted**: a *numeric* vector (in the same order as `truth`) of class predictions



--------------------



## Functionality
### Training Data

Training data objects created using `create_train()` from an `data.frame`
object.
The `create_train()` function can subset the data based on the i
information passed in the `...` (passed to `dplyr::filter()` and are labeled 
by the `group.var =` argument.
A "Response" column to the end of the data frame corresponding to the classes.
The negative and positive classes are the grouped according to this column
and a `dplyr::group_by()` call, creating a "grouped tibble" object.
For example: consider the simple case where groupings are defined by a 
single meta data field called `Species`:

```{r train}
add_noise <- function(x) x + rnorm(length(x), sd = 3)
tr <- tr_iris
class(tr)
```

As is typical with `dplyr::filter()`, multiple conditions can be added
as desired:

```{r train2, eval = FALSE}
tr2 <- create_train(iris, Species == "setosa", Sepal.Length > 5,
                    group.var = Species)
```


-------------------


## Univariate: The `calc.*` Family

The `calc.*` family are a series of wrappers for creating univariate tables
of statistical test results ordered by p-value. 
These functions perform a univariate test on each feature (SOMAmer analyte)
in the `tr_data` object using the existing "Response" column by default
(this can be otherwise defined by the user). 

The resulting `stat.table` object is sorted by *p-value*, and columns
for multiple testing corrections (fdr, Bonferroni) and numeric rank
are added. 

The current suite of univariate tests are:

* `calc.ks()`: KS table
* `calc.t()`: t-test table (paired or unpaired)
* `calc.wilcox()`, `calc.mw()`: Wilcoxon table; Mann-Whitney table
* `calc.cor()`: Correlation table (Spearman, Pearson, Kendall)
* `calc.lr()`: Log-ratio table
* `calc.kw()`: Kruskall-Wallace table
* `calc.lm()`: Linear regression table
* `calc.glm()`: Generalized linear regression table
* `calc.robust.lm()`: Robust linear regression table
* `calc.mackwolfe()`: Mach-Wolfe test table

Some tests, like `calc.t()`, have support for paired analyses. To perform paired
tests the `tr_data` object must be sorted by the "Response" column and then 
by `SampleId` (identifying the pairs) to ensure the samples are paired properly.


#### Examples

```{r calc_t}
splyr:::sim_test_data |>   # from `splyr`
  log10() |>     # log-transform
  calc.t(response = "class_response")   # binary grouping variable
```


---------------------


## Model Fitting
### Splitting Train vs Test Sets

A *very* convenient method to split a data set into training and test sets is:

```{r train_test}
# Add a sample identifier variable for the anti_join()
tr <- dplyr::mutate(tr, id = dplyr::row_number())

train <- tr |>
  dplyr::sample_frac(size = 0.5)        # random selection of rows @ 50%

test <- tr |>
  dplyr::anti_join(train, by = "id") |> # use anti_join to get the sample setdiff
  dplyr::select(-id)                    # remove identifier

train <- dplyr::select(train, -id)      # remove identifier
```


### Robust Naive Bayes

This function fits a naive Bayes model using *robustly* calculated parameters,
i.e. calculated using a least-squares fit to a Gaussian cumulative distribution
function rather than simply `stats::mean()` and `stats::sd()`. This engine
comes from `SomaGlobals::calcRobustGaussFit()`, with starting parameter values
calculated via `stats::median()` and `stats::mad()` passed to `stats::nls()`.

```{r naiveBayes}
nb <- fit_nb(Species ~ ., data = train)
nb
```

#### Plotting a Bivariate Naive Bayes Boundary

If you would like to see how a naive Bayes boundary looks in 2 dimensions:
```{r BayesBoundary, fig.width = 7, fig.height = 7}
data.frame(F1    = tr$Petal.Length,
           F2    = tr$Sepal.Length,
           class = tr$Species) |>
  plotBivariateBayesBoundary(pos.class = "virginica")
```



### Random Forest

Random forest models are often used in SomaLogic analysis, and there are
numerous functional tools to investigate and interrogate these models,
e.g. `backwardsRandomForest()` and `getGini()`:

```{r randomForest}
rf <- randomForest(Species ~ ., data = train)
getGini(rf)
```



----------------------



## Model Performance

As one might expect, there are numerous functions for evaluating model
performance in [`libml`](../index.html). Many S3 methods are provided
to the S3 generic `globalr::calcPredictions`, e.g. `calcPredictions.glm()`.
The following are a few of the standard metrics:

### Confusion Matrix

```{r conf_mat}
pred_nb <- calcPredictions(nb, test)
pred_nb

calc_confusion(test$Species, pred_nb$prob_virginica,
               pos.class = "virginica", cutoff = 0.5)

pred_rf <- calcPredictions(rf, test)
pred_rf

calc_confusion(test$Species, pred_rf$prob_virginica,
               pos.class = "virginica", cutoff = 0.5)
```


### Performance Stats

Typical performance statistics can then be generated by calling the S3
`summary()` generic for the `confusion.matrix` class:

```{r performance}
calc_confusion(test$Species, pred_rf$prob_virginica,
               pos.class = "virginica", cutoff = 0.5) |> summary()
```



### ROC

The primary function for creating ROC curves is called `plotEmpROC()`, or
plot empirical ROC curve. It can be called by providing both the `truth =` and
`predicted =` argument vectors, as well as a `cutoff` for delineating the 
class labels (i.e. operating point default `= 0.5`).

There is a confidence interval box drawn around the operating point that can 
be turned off using the `boxes =` argument.

The `plotEmpROC()` function has support for adding multiple ROC curves to one
plot using the `add =` argument. When this argument is set to `FALSE` (or 0)
a new plot is created; however, when `add = TRUE` (or 1) a second ROC curve 
is added and its corresponding AUC value is placed beneath the AUC for 
the first ROC curve.

When adding a multiple ROC curves, set `add =` to integers `> 1` and each
successive AUC value will be placed below the previous. 
The position and spacing of these AUC label can be adjusted using
the `auc.shift =` argument.

A smoothed curve fit to the empirical ROC can be added using `plot.fit = TRUE`.

```{r roc, fig.width = 7, fig.height = 7}
plotEmpROC(test$Species, pred_nb$prob_virginica, pos.class = "virginica",
           col = "blue")
plotEmpROC(test$Species, pred_rf$prob_virginica, pos.class = "virginica", 
           col = "red", plot.fit = TRUE)
```
