---
title: "Plotting ROC Curves"
output: rmarkdown::html_vignette
author: Amanda Hiser
vignette: >
  %\VignetteIndexEntry{Plotting ROC Curves}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
library(libml)
library(ggplot2)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 5
)
options(width = 100)
```

## Introduction & Brief Review of ROC Curves

This vignette describes how to plot ROC curves using functions found in
**libml**. Before starting on the code walk through, however, we will provide
a quick refresher on ROC curves. Skip to the next section if this information
is already familiar to you.

A **receiver operating characteristic curve**, or **ROC curve**, is a
graphical device that helps to visualize the discriminatory ability of a
binary classifier system. 

To produce a ROC curve, the sensitivity (aka true positive rate) 
and specificity (aka false positive rate) are calculated for different
values from a continuous set of data. To translate this into a visualization,
the ROC curve is made by plotting *sensitivity* (on the y-axis) against
*1–specificity* (on the x-axis) for the calculated values.

A ROC curve that follows the diagonal (unit) line *y = x* produces false
positive results at the same rate as true positive results.
Classifiers with reasonable accuracy should have a ROC curve with its
highest point in the upper left corner of a plot, above 
the *y = x* line. This accuracy can be quantified using the area under
the ROC curve (AUC) metric, which is a measure of the ability of a test
to discriminate between groups or correctly identify when a condition is present.
The higher the AUC, the better the model is at making accurate predictions.
For example, an AUC of 0.5 represents a test with no discriminating ability
(basically no better than chance), while an AUC of 1.0 represents a 
test with perfect discrimination. It is common to see a graphical representations 
of a ROC curve paired with the corresponding numeric AUC value.

**Fun fact:** The ROC curve was first developed by electrical engineers and
radar engineers during World War II, for the purpose of detecting enemy
objects in battlefields by operators of military radar receivers.
This is how the curve got its name.


## Plotting ROC Curves

ROC curves are a frequently used visualization tool for model 
assessment, classifier performance, biomarker evaluation, and more.
Consequently, the **libml** package is well-equipped to easily
and quickly generate and visualize ROC curves and AUC values. 

At the most basic level, individual ROC curves can be generated using the
`geom_roc()` function. This function creates a
[ggplot2](https://ggplot2.tidyverse.org)-compatible 
*"geom"* layer that generates and plots a ROC curve, utilizing the
`ggplot2`-style grammar of graphics. 

The primary input for `geom_roc()` is the output of `roc_xy()` (also from 
**libml**), and this is generally used in support of the wrapper functions 
`plot_emp_roc()` and `plot_boot_roc()`. These functions generate ROC curve plots, 
but provide more advanced options for plot annotation and confidence interval 
generation, respectively.

To illustrate how `geom_roc()` can be used to create a ROC curve, we must
first generate an example data set to work with. The `roc_xy()` function
will then create the `x` and `y` values required to plot the curve
from this data set.

```{r example-data}
# Generate example data
true <- rep(c("control", "disease"), each = 10)
pred <- withr::with_seed(8, c(rnorm(10, mean = 0.4, sd = 0.2),
                              rnorm(10, mean = 0.6, sd = 0.2)))

# Calculate x & y coords for the ROC curve using data generated above
rocxy <- roc_xy(true, pred, "disease") |>
  data.frame()     # must cast to a data frame for ggplot()
head(rocxy)
```

Now that we have obtained the coordinates for the ROC curve,
we can use `ggplot()` and `geom_roc()` to plot the curve:

```{r roc-plot}
# Set up ggplot2-style plot mappings
ggplot(data = rocxy, mapping = aes(x = x, y = y)) +
  geom_roc(color = "blue")    # add curve layer to plot
```

You'll notice that this is a very simplistic plot; because `geom_roc()`
merely creates the ROC curve layer
(without modifying the axis labels, plot title, etc.), the plot above 
is pretty basic and uses the default ggplot2 theme. We can class it up a
bit by applying and/or modifying additional theme elements:

```{r cleaned-plot}
p <- ggplot(data = rocxy, aes(x = x, y = y)) +
  labs(x = "1 - Specificity", y = "Sensitivity") +
  theme(panel.grid = element_blank()) +
  geom_abline(slope = 1, intercept = 0, size = 0.2,
              linetype = "dashed", color = "grey") +
  theme(panel.grid.minor = element_line(linetype = "dashed",
                                        color = "grey", size = 0.2),
        panel.grid.major = element_line(linetype = "dashed",
                                        color = "grey", size = 0.2)) +
  geom_roc(color = "blue")

p
```

A fitted line can be layered on top of this curve using `geom_rocfit()`.
Like other `geom`s, features of the fit line (like color, size, line type, etc.)
can be customized:

```{r fit-curve}
p + geom_rocfit(data = rocxy, color = "blue", lty = "dashed")
```

You'll notice that it took a decent number of lines of code to get to this
point, even though the final product is still a rather simplistic plot.
The wrapper functions mentioned previously alleviate this issue by baking
all of these steps into one single function.


## Convenient Wrappers
### `plot_emp_roc()`

It's often advantageous to annotate ROC curves with AUC values, draw points
along the curve, and add other visual features to make the plot easier
to interpret and more informative.
Additionally, when comparing multiple ROC curves, it's useful to layer the
curves on top of each other to aid in visual comparison and simplify the
overall presentation.

However, adding annotation and/or text layers and customizing theme
elements can become labor intensive and messy when there are 2+ curves
(plus fit lines, in some cases) to plot.
The wrapper function `plot_emp_roc()` streamlines this process by
generating custom annotations under the hood.

```{r single-plot}
plot_emp_roc(truth     = true,
             predicted = pred,
             pos_class = "disease",
             plot_fit  = "both",  # both ROC and fit line should be drawn
             col       = "red",
             shape     = 10)
```

The plots generated by `plot_emp_roc()` can also be layered, to compare
ROC curves generated from various data sets:

```{r layered-plots}
# Generate additional datasets
n <- 50
data <- lapply(1:4, function(.x) {
  withr::with_seed(101,
    data.frame(truth = rep(c("control", "disease"), each = n),
               pred  = c(rnorm(n), rnorm(n, mean = .x / 2)))
  )
}) |> setNames(paste0("df_", 1:4L))

# Create a color palette to distinguish the curves
cols <- unlist(libml:::col_palette)

# Create a base plot & layer additional curves layered on top
plot_emp_roc(data$df_1$truth, data$df_1$pred, "disease", col = cols[1L]) +
  plot_emp_roc(data$df_2$truth, data$df_2$pred, "disease", col = cols[2L], add = 1) +
  plot_emp_roc(data$df_3$truth, data$df_3$pred, "disease", col = cols[3L], add = 2) +
  plot_emp_roc(data$df_4$truth, data$df_4$pred, "disease", col = cols[4L], add = 3)
```

**Note:** when adding plots together, the `add =` argument (integer) *must* be
passed and set to the plotting layer (zero indexed) for the desired curve.
This is used to stagger the AUC text annotation values so that they do not
overlap.


--------------------


### `plot_boot_roc()`

This system/syntax can also be used for the other wrapper function,
`plot_boot_roc()`, which has a similar syntax to `plot_emp_roc()` and is also
built around `geom_roc()`. 
The primary difference between these two is that `plot_emp_roc()` has more
options for customizing the plot appearance, while `plot_boot_roc()` is
designed to generate and visualize bootstrapped CI95 intervals via a shaded
region in addition to the ROC curve.

```{r boot-roc}
# Simulate a new example dataset for this example
n <- 50
true <- rep(c("control", "disease"), each = n)
pred <- withr::with_seed(1, c(rnorm(n, 0.2, 0.2), rnorm(n, 0.5, 0.2)))
plot_boot_roc(true, pred, "disease", shade.color = "blue", color = "red", nboot = 200)
```


These plots can also be layered, with similar syntax to `plot_emp_roc()`:

```{r layered-boot-cI95}
plot_boot_roc(data$df_3$truth, data$df_3$pred, "disease", shade.color = cols[4L],
              color = cols[4L], nboot = 200) +
  plot_boot_roc(data$df_2$truth, data$df_2$pred, "disease", shade.color = cols[3L],
                color = cols[3L], nboot = 200, add = TRUE) +
  plot_boot_roc(data$df_1$truth, data$df_1$pred, "disease", shade.color = cols[2L],
                color = cols[2L], nboot = 200, add = TRUE)
```
