---
title: "Using Nested Tibbles for Univariate Analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using Nested Tibbles for Univariate Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
library(splyr)
library(tidyr)
library(purrr)
library(ggplot2)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
options(width = 100)
```


## Introduction to List-Columns

This vignette utilizes one of the most esoteric concepts in the 
[tidyverse](https://www.tidyverse.org) list columns.
The [tidyverse](https://www.tidyverse.org) is based on tidy data, which 
is based on tables. 
Tables simplify data analysis by making data easy to use, just like Arabic 
numerals simplify math by making numbers easy to use. 
But you may not know that tables can make other things easy to use as well. 
You can use tables to store models, lists, and other tables in a list column. 
The table will store the products of your data analysis in an organized way, 
and you can manipulate the table with other familiar
[tidyverse](https://www.tidyverse.org) tools such as
[dplyr](https://dplyr.tidyverse.org) and [purrr](https://purrr.tidyverse.org).
In fact, they are *designed* work together.
This vignette will look at how you can make list columns and use them in 
a common use-case example.

* A great RStudio webinar from Garrett Grolemund can be found here:
  [how to work with list columns](https://rstudio.com/resources/webinars/how-to-work-with-list-columns/)
* As well as a nice tutorial on how list columns incorporate with the 
  [purrr](https://purrr.tidyverse.org) package by Jenny Bryan can be found here:
  [purrr and list-columns](https://jennybc.github.io/purrr-tutorial/ls13_list-columns.html)


## Example Analysis

A very common task in proteomics is a univariate analysis of RFU based
analyte measurements. This means that each analyte/feature is analyzed
independently and the exact same statistical test is applied repeatedly
over the columns of feature data.
This is so common it forms the basis of most preliminary analyses.

As with most analyses, start by pre-processing the data via 3 steps:

* log10-transform
* center to zero mean
* scale to unit variance

```{r preprocess}
# pre-process simulated data
data  <- splyr::sim_test_data
feats <- grep("^seq.", names(data), value = TRUE)
for ( i in feats ) {
  data[[i]] <- log10(data[[i]])
}
data  <- center_scale(data)
```

Next, initiate a table ([tibble](https://tibble.tidyverse.org)) of analyte
annotations. This will be the starting point of the ever-growing 
analysis "object":

```{r tbl}
t_tbl <- attr(data, "Col.Meta") |>
  dplyr::filter(Units == "RFU") |> # optionally filter features here
  dplyr::mutate(feature = libml:::seqid2apt(SeqId)) |>
  select(feature, SeqId, TargetFullName, EntrezGeneSymbol, UniProt)

head(t_tbl)
```


### Calculate *t*-tests

With the above preparation steps in place, iterate over the rows and add 
columns as the analysis grows, with a combination of `dplyr::mutate()`
and `purrr::map()`:

```{r t-test}
t_tbl <- t_tbl |>
  mutate(
    t_stat  = map(feature, ~ as.formula(paste(.x, "~ class_response")) |>
                  t.test(data = data)),        # fit t-tests
    p_value = map_dbl(t_stat, "p.value"),      # pull out p-values
    fdr     = p.adjust(p_value, method = "fdr") # FDR test correction
  ) |>
  arrange(p_value) |>           # Re-order by `p-value`
  mutate(rank = row_number())   # add ranks column

# each row represents an analyte
length(feats)
nrow(t_tbl)

# the analysis table
t_tbl
```

Done! We now have a self-contained [tibble](https://tibble.tidyverse.org) 
object with the entire t-test
analysis. The `t_stat` list-column contains the "t-tests" and the neighboring
columns contain the corresponding *p*-values and fdr corrections.
The  rows (analytes) are ordered by *p*-value and appropriate rank order has
been added. This would have previously been achieved via `calc.t()`, 
but this is much more self-contained and has a direct link to the target 
annotation data, which is quite useful as we will see below.


-------------


### Plotting in [ggplot2](https://ggplot2.tidyverse.org)

Plotting can be achieved in three basic steps:

* Set up a target map which maps `SeqIds` -> `Targets` (protein names)
  + i.e. for plot titles
* Set up plotting data table
* Plot with [ggplot2](https://ggplot2.tidyverse.org)

#### Dictionary map of feature -> protein names
```{r target-map}
target_map <- head(t_tbl, 9L) |>              # mapping table
  select(feature, Target = TargetFullName)    # feature -> Target (& rename)

target_map
```

#### reshape data into long-format for plotting

```{r plot-data, fig.height = 5, fig.width = 9}
plot_tbl <- undo_center_scale(data) |>                 # log-space; no center-scale
  select(class_response, all_of(target_map$feature)) |># top 9 analytes
  pivot_longer(cols = -class_response, names_to = "feature", values_to = "RFU") |>
  left_join(target_map) |>
  # order factor levels by 't_tbl' rank to order plots below
  mutate(Target = factor(Target, levels = target_map$Target))

plot_tbl
```

#### plot via `ggplot2`

```{r gg-boxplots, fig.height = 5, fig.width = 9}
plot_tbl |>
  ggplot(aes(x = class_response, y = RFU, fill = class_response)) +
  geom_boxplot(alpha = 0.5, outlier.shape = NA) +
  scale_fill_manual(values = c("#24135F", "#00A499")) +
  geom_jitter(shape = 16, width = 0.1, alpha = 0.5) +
  facet_wrap(~ Target) +
  ggtitle("Boxplots of Top 9 Analytes by t-test") +
  labs(y = "log10(RFU)")
```

